Memories of migrations past: Sociality and cognition in dynamic, seasonal environments

### 2021-07-20 text/SocialMemoryMigration_FEE.tex
pushy pushy

@@

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{Eliezer Gurarie\,$^{1,2,*}$, 
	Sriya Potluri\,$^1$, 
	G.~Christopher Cosner\,$^3$,
	R.~Stephen Cantrell\,$^3$,
	William F.~Fagan\,$^1$
}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
	$^{2}$Department of Environmental and Forest Biology, Syracyse, NY   \\
	$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eliezer Gurarie}

\def\corrEmail{egurarie@umd.edu}

@@

		%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
		\section{}
		Seasonal migrations are a widespread and broadly successful strategy for animals to exploit

		\tiny
		\keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
	\end{abstract}

@@

	Seasonal migrations are widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}. Essentially, the evolutionary stability of a migratory strategy relies on the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction outweighting the energetic and survival related costs of migration \citep{Avgar2014}.

	Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017, Merkle2019, Bauer2020}.

	Migratory species may be more vulnerable to environmental change as disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, migratory species might be more resilient to large-scale disruptions due to their wide-ranging mobility \citep{Robinson2009}. A long-distance migration can not rely entirely on local cues, and is therefore only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change.While migration can clearly be a highly successful strategy, the resilience of migratory populations and migratory behavior is an open question and depends on the plasticity and adaptability of a population \citep{Xu2021}.  That plasticity can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citet{Gurarie2017,Xu2021}.

	Cognitive processes, in particular spatial memory, have been shown and argued to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}. Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2020,Fagan2011,Jesmer2018,Berdahl2018}.  However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment is largely unstudied.  Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

@@

	\section{Methods}

	\subsection{Memory movement model}

	In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) a long-term (or \emph{reference}) memory of large-scale movement behavior, (4) a short-term (or \emph{working}) memory that updates movement behavior based on recent experience, and (5) some social aspect to the learned behavior.

@@

	\begin{equation}\label{eq_mainmodel}
		{\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
		\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
		\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + 
		v_m(t)
	\end{equation}

@@

	The second term represents the attraction to a dynamic resource $h$, with the proportionality of the advection to the gradient of the resource given by the parameter $\alpha$ (note, the population and resource distributions are functions of both space and time $u(x,t)$ and $h(x,t)$ - we omit the dependent variables in the notation for brevity). This is the well-studied standard chemotaxic resource-following behavior. We borrow the general notation from our earlier related work \citep{Fagan2017, Fagan2019}.

	The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$. If this function takes the form of a convolution around a non-local kernel $k$, i.e.\textasciitilde{} $v_s(u) = k(x) * u(x)$, and if that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}. We use the kernel analyzed by \citet{Mogilner1999}: $k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$ The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \widehat{u}$, and in a negative direction when $x > \widehat{u}$. The parameter $\lambda$ is a length scale of sociality - roughly one-half the size of the ``swarm'', where $\beta$ is a parameter that quantifies the overall strength of sociality.

	Finally, the last term captures the direct advection that emerges from a memory-driven migratory behavior. This term evolves with a set of parameters $\theta_y$ that slowly change each year $y \in \{0,1,2,...\}$, i.e.~the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.

	The migration speed is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (nominally, winter and summer) $t_1$, $\Delta t_1$, $t_2$, $\Delta t_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$. Thus the remembered migratory speed term is a simple step function given by:

@@

	\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, leave at $t = t_1 + \Delta t_1$, arrive at $x_2$ at $t = t_2$, i.e.~$s_{12} = \frac{x_2-x_1}{t_2 - (t_1 + \Delta t_1)}$ and $s_{21} = \frac{x_1-x_2}{t_1 - (t_2 - \tau + \Delta t_2)}$. This step-like migration function is a one-dimensional version of the migration parameters estimated in empirical studies for individuals \citep{Gurarie2017} and, more relevantly, for populations \citep{Gurarie2019} in empirical studies.

@@

	$$\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y}$$

	\noindent where each of the six parameters is updated identically. The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, \Delta t'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$). The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year. When $\kappa = 1$, the actionable memory is entirely the reference memory.

	The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries i.e.~${\partial u(-\chi,t)}/{\partial t} = {\partial u(\chi,t)}/{\partial t} =0$. As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1. Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process. Our interest is entirely in the ability of a fixed set of memory and movement parameters to navigate an intra- and interannually dynamic, seasonal environment.

@@

	We solved the model numerically on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.\textasciitilde years of 100 days). We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A ``non-surfable'' resource (\emph{island resource}), and weakly surfable resource (\emph{drifting resource}). Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and -30 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively. The island resource is simply two uncorrelated bivariate normal
	distributions 
	$$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$ 
	\noindent where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.~$\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1$.

@@

	Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$. However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse. Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses. These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate. The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.

@@

	The main metrics we are interested in are \emph{migration mismatch}, \emph{foraging efficiency} and \emph{adaptation to trends}.

	Migration mismatch captures the similarity between the migration phenology and the resource phenology. To do this, we compute the total mismatch, i.e.~is a sum of the difference between the migration coefficients and the resource peaks. Spatial mismatch $\text{MM}_x$ is given by the absolute difference between the migration targets and the resource peaks: $\text{MM}_x = |x_1 - m_x| + |x_2 + m_x|$. Temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$\text{MM}_t = max\{t_1 - m_1, m_1 - (t_1 + \Delta t_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + \Delta t_2), 0\}$. Finally, the total mismatch is the sum of these: $\text{TM} = \text{MM}_x + \text{MM}_t$. A mismatch of less than 1 is essentially perfect, we consider a mismatch of 1-5 to be ``good'', and beyond 50 the system can be said to have failed to keep track of the resource dynamics.

	To quantify the foraging efficiency, i.e. the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions. We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is:
	$$\text{FE} = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
	\noindent where the spatial integral is taken over the domain. This metric is constrained to be between 0 and 1.

	For simulations with a constant resource, we ran the model until a quasi-equilibrium (stationary) state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of $0.9999$. Once stationarity was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.

	For numerical runs with climate change, we first run a simulation with a given parameter set until stationarity, as above, and then begin shifting the location of the resource poleward with a slow, moderate or rapid trend ($\gamma_x = 0.25, 0.5$ and 1, respectively), and / or by adding stochasticity (spatial standard deviation 3, 6, 9 or 12). For stochasticity analyses, we compare foraging efficiency across a range of the reference memory parameter $\kappa$.  For analyses that include trend (with or without stochasticity), we quantified the ability of the system to keep track of climate change with a \emph{spatial adaptation} (SA) index. This index is the ratio of the slope of the memory-based migration location over time (i.e.~the fitted slope coefficient of $m_{x,i} = \widehat{\gamma_x} i + m_{x,0}$, where $i$ is the year) against the resource drift parameter $\gamma_x$, i.e.~$\text{SA} = \widehat{\gamma}/\gamma_x$. An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicates that the process is not responding at all to climate change.  Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of migration behavior.

@@

	We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions: (1) Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation? (2) Can this model acquire a migratory behavior from a non-migratory initial condition? (3) What is the role of a reference memory for dealing with stochastic resource dynamics? (4) Can this model adapt when the resource peaks shifts in space?

@@

		\centering
		\begin{tabular}{ll|l}
			\hline\hline
			\multicolumn{3}{l}{\textbf{Memory migration model}}\\ 
			& $\epsilon$ & Diffusion \\ 
			& $\alpha$ & Forage following \\ 
			& $\beta$ & Strength of sociality \\ 
			& $\lambda$ & Spatial scale of sociality \\ 
			& $\kappa$ & Proportion of reference versus working memory \\ 
			& $x_1$, $x_2$ & location of population centroids in summer and winter \\ 
			& $t_1$, $\Delta t_1$ & start and duration of summer season \\ 
			& $t_2$, $\Delta t_2$ & start and duration of winter season \\ 
			& $\kappa$ & proportion memory allocated to reference \\
			&& (long-term) memory vs. working (short-term) memory \\ 
			\multicolumn{3}{l}{\textbf{Resource dynamics}} \\ 
			& $\tau$ & duration of period (year) \\ 
			& $m_x$, $-m_x$ & spatial coordinate of resource peak for summer and winter \\ 
			& $m_t$, $\tau - m_t$ & timing of resource peak for the summer and winter \\ 
			& $\sigma_x$, $\sigma_t$ & time duration and and spatial scale of resource pulse \\ 
			& $\gamma_x$, $\gamma_t$ & rate of change of peak location and timing of resource \\ 
			& $\psi_x$, $\psi_t$ & standard deviation of peak location and timing \\ 
			\multicolumn{3}{l}{\textbf{Metrics}}\\ 
			& $\text{MM}_x$ & spatial migration mismatch \\ 
			& $\text{MM}_t$ & temporal migration mismatch \\ 
			& $\text{TM}$ & total mismatch \\ 
			& $\text{FE}$ & foraging efficiency \\ 
			& $\text{SA}$ & spatial adaptation index \\ 
			\hline\hline
		\end{tabular}
	\end{table}

@@

	\subsection{Adaptation to resource phenology}

>>>>

\usepackage{braket}

@@

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{Eliezer Gurarie\,$^{1,2,*}$, 
	Sriya Potluri\,$^1$, 
	G.~Christopher Cosner\,$^3$,
	R.~Stephen Cantrell\,$^3$,
	William F.~Fagan\,$^1$
}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
	$^{2}$Department of Environmental and Forest Biology, SUNY College of Environmental Science and Forestry, Syracyse, NY 13210, USA.\\
	$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eliezer Gurarie}
\def\corrEmail{egurarie@umd.edu}

@@

		%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
	Seasonal migrations are a widespread and broadly successful strategy for animals to exploit periodic and localized resources over large spatial scales.  It remains an open and largely case-specific question whether long-distance migrations are resilient to environmental disruptions. High levels of mobility suggest an ability to shift ranges that can confer resilience. On the the other hand, a conservative, hard-wired commitment to a risky behavior can be costly if conditions change.

	Mechanisms that contribute to migration include identification and responsiveness to resources, sociality, and cognitive processes spatial memory and learning.  Our goal was to explore the extent to which these factors interact not only to maintain a migratory behavior but also to provide resilience against environmental changes. We develop a relatively simple diffusion-advection model of animal movement in which an endogenous migratory behavior is modified by recent experiences via a memory process, and animals have a social swarming-like behavior over a range of spatial scales.

	We found that this relatively simple framework was able to adapt to a stable, seasonal resource dynamic under a fairly broad range of parameter values. Furthermore, the model was able to acquire an adaptive migration behavior with time. However, the resilience of the process depended on all the parameters under consideration, with many complex trade-offs. For example, the spatial scale of sociality needed to be large enough to capture changes in the resource, but not so large that the acquired collective information was overly diluted.   A long-term reference memory was important for hedging against a highly stochastic process, but a higher weighting of more recent memory was needed for adapting to directional changes in resource phenology.

	Our model provides a general and versatile framework for exploring the interaction of memory, movement, social and resource dynamics, even as environmental conditions globally are undergoing rapid change.

		\tiny
		\keyFont{ \section{Keywords:} Sociality, seasonal migration, collective knowledge, diffusion-advection models, spatial memory} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
	\end{abstract}

@@

	Seasonal migrations are widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}. The evolutionary stability of a migratory strategy essentially relies on the fitness benefits of accessing seasonal resources, whether for energetic gain, predator avoidance, or a suitable  environment for reproduction, outweighing the energetic and survival related costs of migration \citep{Avgar2014}.

	Proximate causes, drivers and mechanisms for migration vary widely across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs and benefits of accessing those highly seasonal and localized resources. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017, Merkle2019, Bauer2020}.

	While migration can be a very successful strategy, with migratory ecotypes of the same species often outnumbering non-migratory conspecifics.  Migratory caribou and reindeer \emph{Rangifer tarandus}, for example, are several orders of magnitude more abundant than non-migratory woodland, mountain and forest ecotypes \citep{FestaBianchet2011,Uboni2016}. However, the question of whether migratory animals are more or less resilient to environmental disruptions in the environment remains open and largely case-specific \citep{Moore2008, HardestyMoore2017, Xu2021}. On the one hand, migratory species may be more vulnerable as disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Seebacher2015, Kauffman2021}. On the other hand, migratory species might be more resilient due to their general wide-ranging mobility \citep{Robinson2009}. The resilience of a migratory populations depends on the plasticity and adaptability of a population, which can take multiple forms, reflecting variation in \textit{where}, \textit{when} and \textit{whether} the migration occurs \citep{Gurarie2017, Xu2021}.

	Cognitive processes, in particular spatial memory, have been shown to be important mechanisms for the reinforcement and maintenance of migration \citep{Merkle2019, Bauer2020}. Similarly, sociality and social learning are likely essential to maintaining migration \citep{Guttal2010, Fagan2011, Jesmer2018, Berdahl2018}.  However, the interacting role of sociality and spatial memory for the plasticity of migration and the resilience of the behavior when faced with a changing environment are generally unknown, though it has been hypothesized that the importance of these cognitive processes depend on the predictability of these resources \citep{RiotteLambert2020}. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

@@

\section{Methods}
~
\subsection{Memory movement model}

	In designing our study, our goal was to develop a minimal heuristic in which the following processes were explicitly modeled: (1) Random or exploratory movement, (2) attraction to resources, (3) sociality in the movements, (4) a long-term (or \emph{reference}) memory of large-scale movement behavior, and (5) a short-term (or \emph{working}) memory that updates movement behavior based on recent experience.

@@

	\begin{equation}\label{eq_mainmodel}
		-{\frac{\partial u}{\partial t}} = -\varepsilon {\frac{\partial^2 u}{\partial x^2}} + 
		\alpha \frac{\partial}{\partial x}\left(u \frac{\partial h}{\partial x}\right) + 
		\beta \frac{\partial}{\partial x}\left(v_s(u)\right) + \frac{\partial}{\partial x}\left(u\,v_m(t)\right)
		\end{equation}

@@

	The second term represents the attraction to a dynamic resource $h$, with the proportionality of the advection to the gradient of the resource given by the parameter $\alpha$ (note, the population and resource distributions are functions of both space and time $u(x,t)$ and $h(x,t)$ - we omit the dependent variables in the notation for brevity). This is the well-studied standard chemotaxic resource-following behavior. We borrow the general notation from earlier related work expanding diffusion-advection models to incorporate non-local information \citep{Fagan2017} and behavioral switching \citep{Fagan2019}.

	The third term captures the collective or social advection term of the population via a non-local, density dependent function $v_s(u,x)$. If this function takes the form of a convolution around a non-local kernel $k$, i.e.~$v_s(u) = k(x) * u(x)$, and if that kernel is odd, an attractive or ``swarming'' behavior can be generated \citep{Mogilner1999}. We use the kernel analyzed by \citet{Mogilner1999}:

	$$k(x) = \frac{x}{2\lambda^2} \exp(-x^2/2\lambda^2).$$

	\noindent The convolution of $u$ with this kernel has the property of pushing the population in a positive direction when $x < \braket{u}$, and in a negative direction when $x > \braket{u}$, where $\braket{u}$ is the mean location of the population. The parameter $\lambda$ is a length scale of sociality, roughly one-half the size of the swarm, and $\beta$ is a parameter that quantifies the overall strength of sociality.

	Finally, the last term captures the direct advection that emerges from a memory-driven migratory behavior. This term evolves with a set of parameters $\theta_y$ that slowly change each year $y \in \{0,1,2,...\}$, i.e.~the count of periods $\tau$: $y = \lfloor t/\tau \rfloor$.  The migration is specified by six parameters $\theta$: the timing of the start and duration of two anticipated seasons (e.g.~summer and winter) $t_1$, $\Delta t_1$, $t_2$, $\Delta t_2$, and the spatial coordinates of the population centroid for each season $x_1$ and $x_2$. The remembered migratory speed term is a simple step function given by:

@@

	\noindent where the migration speeds $s_{12}$ and $s_{21}$ from the respective ranges are set such that they arrive at $x_1$ at $t_1$, depart at $t = t_1 + \Delta t_1$, arrive at $x_2$ at $t = t_2$, and depart at $t_2 + \Delta t_2$. Thus, $s_{12} = (x_2-x_1)/ (t_2 - (t_1 + \Delta t_1))$ and $s_{21} = (x_1-x_2)/(t_1 - (t_2 - \tau + \Delta t_2))$. This step-like migration function is a one-dimensional version of the migration parameters estimated for individuals \citep{Gurarie2017} and populations \citep{Gurarie2019} in empirical studies.

@@

	\begin{equation}\label{eq_memory}
	\theta_{y+1} = \kappa^y \, \theta_o + \left(1-\kappa^y\right)\,\widehat{\theta_y} 
\end{equation}

	\noindent where each of the six parameters is updated according to equaton \ref{eq_memory} identically. The estimates $\widehat{\theta_y}$ are obtained via a least-squares minimization of the migration track ($m(t,\theta) = \int_0^t v_m(t',\theta_y) \, \Delta t'$) against the spatial mean of the population process in year $y$ (i.e.~$\widehat{u}(t) = \int_X u_y(t, x) dx$). The parameter $\kappa \in (0,1)$ captures the reliance on that long-term memory. When $\kappa = 0$, all of the actionable memory is from the preceding year. When $\kappa = 1$, the actionable memory is entirely the reference memory.

	The model is confined to a one-dimensional bounded domain $[-\chi,\chi]$, with no flux outside of the boundaries. Formally, this no-flux condition means the following conditions must be met

	$$
	\begin{cases}
		\varepsilon \frac{\partial u}{\partial x} - 		
	\alpha \left(u \frac{\partial h}{\partial x}\right) - \beta(v_x(u)) - (uv_m(t)) = 0 & \text{where}\,\, x = \chi\\
	-\varepsilon \frac{\partial u}{\partial x} - 
	\alpha \left(u \frac{\partial h}{\partial x}\right) - \beta(v_x(u)) - (uv_m(t)) = 0 & \text{where}\,\, x = -\chi
	\end{cases}
$$

	\noindent In practice, the design of our resource space (see below) and other parameterization lead 0 or near 0 values of both $h(x)$ and $u(x)$, and the simpler ${\partial u(-\chi,t)}/{\partial t} = {\partial u(\chi,t)}/{\partial t} =0$ boundary condition provides a good approximation.

	As there are no birth or death processes, the total population remains fixed and constant, for convenience integrating to 1. Furthermore, the parameters remain constant throughout time, with no adaptation or mutation-selection process. Our interest is in the ability of a fixed set of movement and memory parameters to navigate an intra- and interannually dynamic, seasonal environment.

@@

	We ran this model on a spatial domain $x \in [-100,100]$, and a periodicity $\tau = 100$ (i.e.~100 day year~s). We were interested in an approximately periodic resource dynamic, i.e.~one in which $h(x,t \approx h(x, t-\tau))$. We generated two types of resource distributions. A ``non-surfable'' resource (\emph{island resource}), and weakly surfable resource (\emph{drifting resource}). Both are characterized by a peak in time and space centered at $m_x$ at $m_t$, and $-m_x$ at $\tau - m_t$ (for example, locations 30 and -30 at times 25 and 75, respectively). These pulses have a shared time scale of duration $s_t$ and a spatial scale of extent $s_x$, the standard deviation in the time and space dimension respectively. The island resource is simply two uncorrelated bivariate normal
	distributions 
	$$h(x,t) = K\,(\Phi(m_x, m_t, s_x, s_t) + \Phi(-m_x, \tau - m_t, s_x, s_t))$$ 
	\noindent where $\Phi$ is the bivariate Gaussian distribution function, and the normalizing constant $K$ is selected such that the average total amount of resource throughout the year is 1, i.e.~$\frac{1}{\tau} \int_T\int_X h(x,t) dx\,dt = 1$.

@@

	Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$. However, in scenarios exploring climate change we allow the peaks to vary with directional trend and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse. Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a shift of the resource towards the extremes of the domain. While we did note explore phenological shifts in timing, those can readily be modeled as well. These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate \citep{Renner2018}. The spatial and temporal scales of the resource peak ($s_x$ and $s_t$) remain constant in all of our simulations.

@@

	The main metrics we were interested in are \emph{migration mismatch}, \emph{foraging efficiency} and \emph{adaptation to directional trends}.

	Migration mismatch captures the combined difference between the migration phenology and the resource phenology in time and space. Spatial mismatch $\text{MM}_x$ is the absolute difference between the migration targets and the resource peaks: $\text{MM}_x = |x_1 - m_x| + |x_2 + m_x|$. Temporal mismatch is the difference between the arrival time and the peak of the resource if arrival is post-peak, the difference between the departure time and the peak of the resource if departure is pre-peak, and 0 if the seasonal duration spans the peak, i.e.~$\text{MM}_t = max\{t_1 - m_1, m_1 - (t_1 + \Delta t_1), 0\} + max\{t_2 - m_2, m_2 - (t_2 + \Delta t_2), 0\}$. Thus, the total mismatch is the sum of these: $\text{TM} = \text{MM}_x + \text{MM}_t$. A mismatch of less than 1 is essentially perfect, a mismatch of 1-5 we consider excellent, and beyond 50 the system can be said to have failed to keep track of the resource dynamics.

	To quantify the foraging efficiency, i.e.~the organisms' ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya coefficient \citep{Bhattacharyya1943} which quantifies the similarity between two distributions. We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is:

	$$\text{FE} = \frac{1}{\tau} \int_{0}^\tau \int_{-\chi}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$

	\noindent where the spatial integral is taken over the domain. This metric is constrained to be between 0 and 1.

	For simulations with a constant resource, we ran the model until a quasi-equilibrium (stationary) state was achieved, i.e.~where the Bhattacharya index of the population distribution across subsequent years reached a value of $0.99999$. Once stationarity was attained, we computed the migration mismatch and foraging efficiency metrics, as well as the number of years required to reach stationarity.

	For numerical runs with climate change, we first run a simulation with a given parameter set until stationarity, as above, and then begin shifting the location of the resource poleward with a slow, moderate or rapid trend ($\gamma_x = 0.25, 0.5$ and 1, respectively), and / or by adding stochasticity (spatial standard deviation 3, 6, 9 or 12). For stochasticity analyses, we compare foraging efficiency across a range of the reference memory parameter $\kappa$.  For analyses that included directional trends, with or without stochasticity, we quantified the ability of the system to keep track of climate change with a \emph{spatial adaptation} (SA) index. This index is the ratio of the slope of the memory-based migration location over time, i.e.~$\text{SA} = \widehat{\gamma}/\gamma_x$  where the adaptation slope estimate is the regression coefficient of the spatial coordinate of the migration against time (i.e.~$m_{x,i} = \widehat{\gamma_x} i + m_{x,0}$, where $i$ is the year), and $\gamma_x$ is the rate of drift of the resource peak (table \ref{table_definitions}). An SA equal to 1 suggests that the process is keeping up with climate change, an SA of 0 indicates that the process is not responding at all to climate change.  Values greater than 1 (super-adaptation) are possible, as are values less than 1, which correspond to a loss of migration behavior.

@@

	We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions: (1) Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation? (2) Can this model acquire a migratory behavior from a non-migratory initial condition? (3) What is the role of a reference memory for dealing with stochastic resource dynamics? (4) Can this model adapt when the resource peaks shifts in space?  Details of parameter combinations and reported metrics are provided in respective results sections.

@@

		\centering
		\begin{tabular}{ll|l}
			\hline\hline
			\multicolumn{3}{l}{\textbf{Memory migration model}}\\ 
			& $\varepsilon$ & Diffusion \\ 
			& $\alpha$ & Strength of resource following \\ 
			& $\beta$ & Strength of sociality \\ 
			& $\lambda$ & Spatial scale of sociality \\ 
			& $\kappa$ & Proportion of reference versus working memory \\ 
			& $x_1$, $x_2$ & location of population centroids in summer and winter \\ 
			& $t_1$, $\Delta t_1$ & start and duration of summer season \\ 
			& $t_2$, $\Delta t_2$ & start and duration of winter season \\ 
			& $\kappa$ & proportion memory allocated to reference \\
			&& (long-term) memory vs. working (short-term) memory \\ 
			\multicolumn{3}{l}{\textbf{Resource dynamics}} \\ 
			& $\tau$ & duration of period (year) \\ 
			& $m_x$, $-m_x$ & spatial coordinate of resource peak for summer and winter \\ 
			& $m_t$, $\tau - m_t$ & timing of resource peak for the summer and winter \\ 
			& $\sigma_x$, $\sigma_t$ & time duration and and spatial scale of resource pulse \\ 
			& $\gamma_x$, $\gamma_t$ & rate of change of peak location and timing of resource \\ 
			& $\psi_x$, $\psi_t$ & standard deviation of peak location and timing \\ 
			\multicolumn{3}{l}{\textbf{Metrics}}\\ 
			& $\text{MM}_x$ & spatial migration mismatch \\ 
			& $\text{MM}_t$ & temporal migration mismatch \\ 
			& $\text{TM}$ & total mismatch \\ 
			& $\text{FE}$ & foraging efficiency \\ 
			& $\text{SA}$ & spatial adaptation index \\ 
			\hline\hline
		\end{tabular}
	\end{table}

@@

\subsection{Adaptation to resource phenology}\label{results_adaptation}
